---
title: "Flu & CV-19 trends"
subtitle: "Figures 3, 4 and supplementary figure S3"
author: "Minttu Rönn"
date: '`r format(Sys.Date(), "%Y-%b-%d")`'
output:
  pdf_document: 
    fig_caption: yes
    includes:
      in_header: 
  html_document:
    keep_md: yes
urlcolor: blue
---

```{r global-options, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(cowplot)
library(data.table)
library(ggplot2)
library(ggpubr)
library(here)
library(lme4)
library(lubridate)
library(RColorBrewer)
library(readr)
library(RSocrata)
library(tidyverse)
library(tidycensus)
library(tigris)

```


```{r}

# Election data
source(here('code/read_electiondata.R'))

# FLU-NIS data
source(here('code/read_pubavaildata_final.R'))

# Material for the NIS-FLU descriptive figures
source(here('code/nisflu_figures_specs.R'))
  
```

\newpage

## FIGURE 3 

State-level correlation between 2020 vote share for Biden in 2020 presidential elections (x-axis) and flu vaccine coverage (y-axis) by age (rows) and by flu season (columns). Washington DC excluded from the figure.

```{r}

flust %>%
  filter(vaccine == "Seasonal Influenza" & dimension_type == "Age") %>%
  filter(year > 2010 & !is.na(state_code) & state_code != "PR" & state_code != "DC" & state_code != "GU") %>%   ## need to remove 
 # left_join(elect, by=c("state_code", "elect_year")) %>%
  dplyr::select(year, est, age_group, state_code)   %>% # est, est_ll, est_ul, 
  pivot_wider(names_from = c(age_group, year), values_from = c(est)) %>%
  dplyr::select(ends_with(".5"), state_code) %>%
  pivot_longer(cols = ends_with(".5"), names_to = "age", values_to = "coverage")  %>%
  mutate(age = factor(age, levels = fct_names)) %>% 
  mutate(year = as.numeric(gsub('.*_', '', age))) %>%
  filter(!is.na(coverage))  -> flust_temp

flust_temp$elect_year <- NA
flust_temp$elect_year[flust_temp$year > 2020] <- 2020
flust_temp$elect_year[flust_temp$year < 2020] <- 2016
flust_temp$elect_year[flust_temp$year < 2016] <- 2012
flust_temp$elect_year[flust_temp$year < 2012] <- 2008

flust_temp %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(d_voteperc = vote_share * 100)  %>%
  ggplot(aes(y=coverage, x=d_voteperc)) +
  geom_point(size = 0.1) +
  geom_smooth(aes(y=coverage, x=d_voteperc), 
               se=TRUE, size=0.5, method = lm) +
  facet_wrap(~age, ncol = 11) + # labeller = labeller(age = label_wrap_gen(width = 15))
  stat_cor(method = "pearson", aes(label = ..r.label..), label.y = 9,  size=2, color = "blue") +
  ylim(c(0, 100)) + xlim(c(10, 80)) +
  min_theme  + 
  ylab( "  65+ years         50-64 years       18-49 years      13-17 years      5-12 years         0.5-4 years  \n") +
  ggtitle(  " 2010-2011         2011-2012      2012-2013        2013-2014       2014-2015      2015-2016        2016-2017        2017-2018        2018-2019        2019-2020        2020-2021  \n")  -> pst4

ggsave( here(paste("output/", 'FIG3-flu-state-age-', Sys.Date(),'.png', sep="")),
        device = png(),
        plot = pst4,
        width =18,
        bg = "white",
        height = 11,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()

pst4

```
\newpage

## FIGURE 4

Predicted and observed flu coverage by age for the 2020-2021 flu season. Correlation between flu vaccine coverage and vote share for Democrats by age

```{r}

# Regression data

flust %>%
  filter(vaccine == "Seasonal Influenza") %>%
  filter(year > 2010 & !is.na(state_code) & state_code != "PR" & state_code != "DC" & state_code != "GU") %>%   
  dplyr::select(year, est, est_ll, est_ul, age_group, dimension_type, dimension, raceth, state_code, population_sample_size)  %>%  
  mutate(est_se = -(est_ll-est)/1.96 ) %>%
  mutate(samplesize = as.numeric(population_sample_size)) %>%
  mutate(age_group = factor(age_group, levels = c("All", "6 Months - 4 Years", "5-12 Years", "13-17 Years", "18-49 Years", "50-64 Years", "65+ Years"))) %>%
  mutate(raceth = factor(raceth, levels = c("All", "White, NH", "AI/AN, NH", "Asian, NH", "Black, NH", "Hispanic", "Other" ))) -> flust_base

  flust_base$elect_year <- NA
  flust_base$elect_year[flust_base$year > 2020] <- 2020
  flust_base$elect_year[flust_base$year < 2020] <- 2016
  flust_base$elect_year[flust_base$year < 2016] <- 2012
  flust_base$elect_year[flust_base$year < 2012] <- 2008
  
  flust_base %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(democrat_voteperc = vote_share * 100) %>%
  filter(!is.na(vote_share))  %>%
  filter(year.x < 2020.5) -> flust_train
    
  flust_base %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(democrat_voteperc = vote_share * 100) %>%
  filter(!is.na(vote_share)) %>%
  filter(year.x == 2020.5) %>%
  rename(est_data = est) -> flust_test

  flust_test$est <- NA
  
```




```{r}

# Regression and prediction

flu.lm5 <- lmer(est ~  year.x + democrat_voteperc +  (1 + age_group | state_code ), data=flust_train) 

flust_test$est_pred <- predict(flu.lm5, newdata=flust_test, type = "response", allow.new.levels = TRUE)


```

```{r}

# SCATTER PLOT - AGE
  flust_test %>%
   filter(age_group!= "All") %>%
   dplyr::select(est_data, democrat_voteperc, est_pred, age_group) %>%
   rename(Observed = est_data) %>%
   rename(Predicted = est_pred) %>%
   pivot_longer(cols = c(Observed, Predicted), names_to = "Estimates", values_to = "estimate") %>%
   ggplot(aes(y=estimate, x=democrat_voteperc, colour=Estimates))+
   geom_point()+
   geom_smooth(method = lm,  size = 0.6) + 
   stat_cor(method = "pearson", aes(label = ..r.label..),  label.x = 0, label.y.npc = "bottom", show.legend = FALSE, size=3)+
    scale_colour_manual(values = c("royalblue", "gray20")) +
    xlab("2020 vote share for Democrats") +
    ylab("Coverage") +
    xlim(c(0, 100)) + ylim(c(0, 100)) +
    facet_wrap(~age_group, ncol = 6) +
    theme_minimal_hgrid() -> pst10

  ggsave( here(paste("output/", 'scatter-age_', Sys.Date(),'.png', sep="")),
        device = png(),
        plot = pst10,
        width =30,
        bg = "white",
        height = 12,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()

```

```{r p0, out.height= "200px"}

p0 <- file.path(here(paste("output/", 'scatter-age_', Sys.Date(),'.png', sep="")))

knitr::include_graphics(p0)

```

\newpage

## FIGURE S3

State level flu vaccination coverage (NIS-FLU) by age (rows) and by HHS region (columns), organized by flu season 2010-2021 (x-axis). HHS regions are ordered by 2020 vote share: Region 1 has the highest share for Democrats and Region 7 has the lowest vote share. The last flu season (2020-2021) is highlighted in orange.

```{r}

flust %>%
  filter(vaccine == "Seasonal Influenza" & dimension_type == "Age") %>%
  left_join(elect, by="state_code") %>%
  filter(year.x > 2010) %>%
  dplyr::select(year.x, est, age_group, state_code, vote_share, geography.y)  %>% # est, est_ll, est_ul, 
  filter(!is.na(vote_share))  %>%
  mutate(binaryvote = as.factor(ifelse(vote_share>0.5, ">50% vote_share", "<=50% vote_share"))) %>%
  mutate(biden_voteperc = vote_share * 100) %>%
  mutate(hhs_region = factor(geography.y, levels = c("Region 1", "Region 9", "Region 2",
                                                     "Region 3", "Region 10", "Region 5",
                                                     "Region 8", "Region 4", "Region 6", 
                                                     "Region 7"))) %>%
  ggplot(aes(y=est, x=year.x, identitity= state_code, color=as.character(year.x))) +
  geom_line(size = 0.12, color="gray20") +
  geom_point(size = 0.3, shape = 20) +
  facet_wrap(~age_group+hhs_region, ncol =10) +
  ylab("Flu vaccine uptake") + xlab("Flu season") +
  scale_color_manual(values = c(rep("gray30", 10), "orange")) + 
  theme_nothing() +
  theme(axis.text.x = element_text(hjust=0, vjust=0.6, size = 6, angle = 90, color = "black", face = "plain"),
        axis.text.y = element_text(hjust=0, vjust=0.6, size = 6, color = "black", face = "plain"),
        strip.text.x = element_text(hjust=0, vjust=1, size = 5, color = "black", face = "plain"),
        panel.spacing.x = unit(0.2, "lines"),
        panel.spacing.y = unit(0.1, "lines"),
        legend.position="none")  -> pst6

ggsave( here(paste("output/", 'FIGS3-flu-state-region-age-time_', Sys.Date(),'.png', sep="")),
        device = png(),
        plot = pst6,
        width =18,
        bg = "white",
        height = 15,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()

pst6

```

