---
title: "Time trends between vaccination coverage and voting patterns before and during the COVID-19 pandemic: analysis of COVID-19 and flu surveys in the United States"
subtitle: "Figures 3, 4 and supplementary figure S3"
author: "Minttu Rönn"
date: '`r format(Sys.Date(), "%Y-%b-%d")`'
output:
  pdf_document: 
    fig_caption: yes
    includes:
      in_header: 
  html_document:
    keep_md: yes
urlcolor: blue
---

```{r global-options, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(cowplot)
library(data.table)
library(ggplot2)
library(ggpubr)
library(here)
library(lme4)
library(lubridate)
library(RColorBrewer)
library(readr)
library(RSocrata)
library(tidyverse)
library(tidycensus)
library(tigris)

```


```{r}

# Election data
source(here('code/read_electiondata.R'))

# FLU-NIS data
source(here('code/read_pubavaildata_final.R'))

# Material for the NIS-FLU descriptive figures
source(here('code/nisflu_figures_specs.R'))


```

\newpage

## FIGURE 2 

State-level correlation between 2020 vote share for Biden in 2020 presidential elections (x-axis) and flu vaccine coverage (y-axis) by age (rows) and by flu season (columns). Washington DC excluded from the figure.

```{r}

flust %>%
  filter(vaccine == "Seasonal Influenza" & dimension_type == "Age") %>%
  filter(year > 2010 & !is.na(state_code) & state_code != "PR" & state_code != "DC" & state_code != "GU") %>%   ## need to remove 
 # left_join(elect, by=c("state_code", "elect_year")) %>%
  dplyr::select(year, est, age_group, state_code) %>% 
  pivot_wider(names_from = c(age_group, year), values_from = c(est)) %>%
  dplyr::select(ends_with(".5"), state_code) %>%
  pivot_longer(cols = ends_with(".5"), names_to = "age", values_to = "coverage")  %>%
  mutate(age = factor(age, levels = fct_names)) %>% 
  mutate(year = as.numeric(gsub('.*_', '', age))) %>%
  filter(!is.na(coverage))  -> flust_temp

flust_temp$elect_year <- NA
flust_temp$elect_year[flust_temp$year > 2020] <- 2020
flust_temp$elect_year[flust_temp$year < 2020] <- 2016
flust_temp$elect_year[flust_temp$year < 2016] <- 2012
flust_temp$elect_year[flust_temp$year < 2012] <- 2008

flust_temp %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(d_voteperc = vote_share * 100)  %>%
  ggplot(aes(y=coverage, x=d_voteperc)) +
  geom_point(size = 0.1, color="gray50") +
  geom_smooth(aes(y=coverage, x=d_voteperc), 
               se=TRUE, size=0.3, method = lm, color = "black") +
  facet_wrap(~age, ncol = 12) + # labeller = labeller(age = label_wrap_gen(width = 15))
  stat_cor(method = "pearson", aes(label = ..r.label..), label.y = 9,  size=2, color = "black") +
  ylim(c(0, 100)) + xlim(c(10, 80)) +
  min_theme  + 
  ylab( "  65+ years         50-64 years       18-49 years      13-17 years      5-12 years         0.5-4 years  \n") +
  ggtitle(  "2010-2011      2011-2012     2012-2013      2013-2014     2014-2015      2015-2016      2016-2017    2017-2018     2018-2019       2019-2020     2020-2021     2021-2022   \n")  -> pst4

ggsave( here(paste("output/", 'FIG-3-', Sys.Date(),'.png', sep="")),
        device = png(),
        plot = pst4,
        width =18,
        bg = "white",
        height = 11,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()

pst4

```
\newpage

## FIGURE 3

Predicted and observed flu coverage by age for the 2020-2021 flu season. Correlation between flu vaccine coverage and vote share for Democrats by age

```{r}

# Regression data
flust %>%
  filter(vaccine == "Seasonal Influenza") %>%
  filter(age_group!= "All") %>%
  filter(year > 2010 & !is.na(state_code) & state_code != "PR" & state_code != "DC" & state_code != "GU") %>%   
  dplyr::select(year, est, est_ll, est_ul, age_group, dimension_type, dimension, raceth, state_code, population_sample_size)  %>%  
  mutate(samplesize = as.numeric(population_sample_size)) %>%
  mutate(age_group = factor(age_group, levels = c("All", "6 Months - 4 Years", "5-12 Years", "13-17 Years", "18-49 Years", "50-64 Years", "65+ Years"))) %>%
  mutate(raceth = factor(raceth, levels = c("All", "White, NH", "AI/AN, NH", "Asian, NH", "Black, NH", "Hispanic", "Other" ))) -> flust_base

  flust_base$elect_year <- NA
  flust_base$elect_year[flust_base$year > 2020] <- 2020
  flust_base$elect_year[flust_base$year < 2020] <- 2016
  flust_base$elect_year[flust_base$year < 2016] <- 2012
  flust_base$elect_year[flust_base$year < 2012] <- 2008
  
  # center year
  flust_base$year_centered <-  flust_base$year - mean(flust_base$year)
  
  # create state+age variable
  flust_base$age_state <- paste0(flust_base$age_group, "_", flust_base$state_code)
 
  flust_base %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(democrat_voteperc = vote_share * 100) %>%
  filter(!is.na(vote_share))  %>%
  filter(year.x < 2020.5) -> flust_train
    
  flust_base %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(democrat_voteperc = vote_share * 100) %>%
  filter(!is.na(vote_share)) %>%
  filter(year.x > 2020) %>%
  rename(est_data = est)-> flust_test

  flust_test$est <- NA

```



```{r}

# Regression and prediction

flu.lm5 <- lmer(est ~  year_centered + democrat_voteperc + state_code + age_group + (1 | age_state  ), data=flust_train) 

# vcv <- vcov(flu.lm5)
# cor_mat <- cov2cor(vcv)
# diag(cor_mat) <- 0
# cor_mat

flust_test$est_pred <- predict(flu.lm5, newdata=flust_test, type = "response", allow.new.levels = TRUE)


```

```{r}

# SCATTER PLOT - AGE
  flust_test %>%
   dplyr::select(est_data, democrat_voteperc, est_pred, age_group, year.x, state_code)  %>%
   mutate(flu_season = ifelse(year.x==2020.5, "2020-2021", "2021-2022")) %>%
   rename(Observed = est_data) %>%
   rename(Predicted = est_pred)  %>%
   pivot_longer(cols = c(Observed, Predicted), names_to = "Estimates", values_to = "estimate") %>%
   ggplot(aes(y=estimate, x=democrat_voteperc, colour=Estimates, fill=Estimates))+
   geom_point(size = 0.6, shape=21)+
   geom_smooth(method = lm,  size = 0.2) + 
   stat_cor(method = "pearson", aes(label = ..r.label..),  label.x = 0, label.y.npc = "bottom", show.legend = FALSE, size=3)+
    scale_shape_manual(values = c(1, 21)) +
    scale_colour_manual(values = c( "grey50", "gray30")) +
    scale_fill_manual(values = c( "grey50", "white")) +
    facet_wrap(~flu_season+age_group, ncol = 6) +
  ylab("Vaccination coverage") + 
  xlab("2020 vote share for the Democratic candidate") + 
  ylim(c(0, 100)) + xlim(c(20, 90)) + 
  min_theme_fig4 -> pst10
# 
#     xlab("0.5-4 years   5-12 years    13-17 years    18-49 years    50-64 years   65+ years") +
#     ylab("       2021-2022              2020-2021") +
#     xlim(c(0, 100)) + ylim(c(0, 100)) +
#     facet_wrap(~flu_season+age_group, ncol = 6) +
#     min_theme_legend   -> pst10

  ggsave( here(paste("output/", 'FIG-3_', Sys.Date(),'.png', sep="")),
        device = png(),
        plot = pst10,
        width =25,
        bg = "white",
        height = 14,
        units = "cm",
        dpi = 500,
        limitsize = TRUE)
dev.off()

```

```{r p0, out.height= "200px"}

p0 <- file.path(here(paste("output/", 'FIG-4_', Sys.Date(),'.png', sep="")))

knitr::include_graphics(p0)

```

\newpage

## EXTRA FIGURE TO MATCH 2

```{r}

flust_overall %>%
  filter(state_code != "PR" & state_code != "DC" & state_code != "GU") %>%
   dplyr::select(est, state_code, year_season) %>%
  mutate(elect_year = 2020)  %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  ggplot(aes(y=est, x=100*vote_share, color=year_season)) +
  geom_point(size=0.3) +
  xlab("2020 vote share for the Democrats") + ylab("Flu vaccination coverage") + #ylim(c(0.5, 2.5))  +
  stat_cor(method = "pearson", show.legend = FALSE, aes(label = ..r.label..), 
           label.x = 0, label.y.npc = "bottom", size=4) +
  xlim(c(0,100))  +  ylim(c(0,100))  + 
  scale_color_manual(values = mycols2[c(1,7)]) +
  geom_smooth(method = lm, size = 0.3) +
   theme_pubclean() +
  theme(axis.text.x = element_text(hjust=0, vjust=0.6, 
                                   size = 8, angle = 0, 
                                   color = "black", face = "plain"),
        axis.text.y = element_text(hjust=1, vjust=0.6, 
                                   size = 8, angle = 0, 
                                   color = "black", face = "plain"),
        legend.position="right")  -> p1

ggsave( here(paste("output/", 'Fig1-C-NIS-fluseason', Sys.Date(), '.png', sep="")),
        device = png(),
        plot = p1,
        width = 15,
        bg = "white",
        height = 10,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()


flust_overall %>%
  filter(state_code != "PR" & state_code != "DC" & state_code != "GU") %>%
   dplyr::select(est, state_code, year_season) %>%
  pivot_wider(names_from=year_season, values_from=est)  %>%
  mutate(elect_year = 2020)  %>%
  left_join(elect, by=c("state_code", "elect_year")) %>%
  mutate(difference = (`2021-22` - `2020-21`)/`2021-22`) %>%
  ggplot(aes(y=100*difference, x=100*vote_share)) +
  geom_point(size=0.3) +
  xlab("2020 vote share for the Democrats") + ylab("Relative difference (%)") + #ylim(c(0.5, 2.5))  +
  stat_cor(method = "pearson", show.legend = FALSE, aes(label = ..r.label..), 
           label.x = 0, label.y.npc = "bottom", size=4) +
  xlim(c(0,100))  +  ylim(c(-20, 20))  + 
  scale_color_manual(values = mycols2[c(8)]) +
  geom_smooth(method = lm, size = 0.3) +
   theme_pubclean() +
  theme(axis.text.x = element_text(hjust=0, vjust=0.6, 
                                   size = 8, angle = 0, 
                                   color = "black", face = "plain"),
        axis.text.y = element_text(hjust=1, vjust=0.6, 
                                   size = 8, angle = 0, 
                                   color = "black", face = "plain"),
        legend.position="right") -> p2

ggsave( here(paste("output/", 'Fig2-D-NIS-fluseason', Sys.Date(), '.png', sep="")),
        device = png(),
        plot = p2,
        width = 15,
        bg = "white",
        height = 10,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()
  

```


\newpage

## FIGURE S3

State level flu vaccination coverage (NIS-FLU) by age (rows) and by HHS region (columns), organized by flu season 2010-2021 (x-axis). HHS regions are ordered by 2020 vote share: Region 1 has the highest share for Democrats and Region 7 has the lowest vote share. The last flu season (2020-2021) is highlighted in orange.

```{r}

flust %>%
  filter(vaccine == "Seasonal Influenza" & dimension_type == "Age") %>%
  left_join(elect, by="state_code") %>%
  filter(year.x > 2010) %>%
  dplyr::select(year.x, est, age_group, state_code, vote_share, geography.y, coverage_estimate)  %>% 
  filter(!is.na(vote_share)) %>%
  filter(coverage_estimate != "NR") %>%    # remove estimates without coverage data (coverage_estimate = NR)
  mutate(binaryvote = as.factor(ifelse(vote_share>0.5, ">50% vote_share", "<=50% vote_share"))) %>%
  mutate(biden_voteperc = vote_share * 100) %>%
  mutate(hhs_region = factor(geography.y, levels = c("Region 1", "Region 9", "Region 2",
                                                     "Region 3", "Region 10", "Region 5",
                                                     "Region 8", "Region 4", "Region 6", 
                                                     "Region 7"))) %>%
  ggplot(aes(y=est, x=year.x, identitity= state_code, color=as.character(year.x))) +
  geom_line(size = 0.12, color="gray20") +
  geom_point(size = 0.3, shape = 20) +
  facet_wrap(~age_group+hhs_region, ncol =10) +
  ylab("Flu vaccine uptake") + xlab("Flu season") +
  scale_color_manual(values = c(rep("gray30", 10), "orange", "orange")) + 
  theme_nothing() +
  theme(axis.text.x = element_text(hjust=0, vjust=0.6, size = 6, angle = 90, color = "black", face = "plain"),
        axis.text.y = element_text(hjust=0, vjust=0.6, size = 6, color = "black", face = "plain"),
        strip.text.x = element_text(hjust=0, vjust=1, size = 5, color = "black", face = "plain"),
        panel.spacing.x = unit(0.5, "lines"),
        panel.spacing.y = unit(0.1, "lines"),
        legend.position="none")  -> pst6

ggsave( here(paste("output/", 'FIG-S3_', Sys.Date(),'.png', sep="")),
        device = png(),
        plot = pst6,
        width =23,
        bg = "white",
        height = 18,
        units = "cm",
        dpi = 300,
        limitsize = TRUE)
dev.off()

pst6

```

