---
title: "Vaccination and voting patterns in the United States: analysis of COVID-19 and flu surveys from 2010 to 2022"
subtitle: "Figures 1 (A,B,C), supplementary figure S1"
author: "Minttu Rönn"
date: '`r format(Sys.Date(), "%Y-%b-%d")`'
output:
  pdf_document: 
    fig_caption: yes
    includes:
      in_header: 
  html_document:
    keep_md: yes
urlcolor: blue
---

```{r global-options, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(cowplot)
library(data.table)
library(ggplot2)
library(ggpubr)
library(here)
library(lme4)
library(lubridate)
library(RColorBrewer)
library(readr)
library(RSocrata)
library(tidyverse)
library(tidycensus)
library(tigris)

```

# Part 1 has code for

- Supplemental Figure S1A, S1B, and S1C: *Comparison of correlation coefficient across state-level COVID-19 vaccination estimates by month.* 
    * ACM data are continuously updated (as are CDC data), and ACM data require a little more checking too see if anything has changed 
- Supplement figure S1: *Absolute monthly difference in COVID-19 vaccination coverage, compared at state-level to previous month.*


```{r}

# Election data
source(here('code/read_electiondata.R'))

# FLU-NIS and NIS-ACM data
source(here('code/read_pubavaildata_final.R'))

mycols2 = c("turquoise4",  "turquoise2",  "springgreen2", "yellowgreen", 
             "wheat2", "orange", "coral2", "red2",  "brown", 
             "violet", "mediumpurple1", "blueviolet", "slateblue", "navyblue", "navy", "brown", "red", "pink", "orange", "orange3", "brown1", "brown3", "steelblue3",    
             "navyblue", "navy")



```

\newpage

## SUPPLEMENTAL FIGURE S1 A 

Comparison of correlation coefficient across state-level COVID-19 vaccination estimates by month. We have omitted Washington D.C. as an outlier.  Pearson correlation coefficient (R) shown by month in panels A,C,D, and as a bar plot in panel B.

```{r}

# CDC Surveillance
# check the max date plotted here

 cv19_bymonth %>%
    mutate(month2 = month(month, label = TRUE, abbr = FALSE)) %>%
    filter(!is.na(vote_share) & date > ymd("2021-01-01") & date < ymd("2022-12-01")) %>%
    filter(state_code != "DC") %>%
    mutate(Year_Month =  format(date, '%Y-%m'))  %>%
   # relocate(Year_Month, .after = date) -> test %>%
    ggplot(aes(x=100*vote_share, y=est, color=Year_Month))  +
    geom_point(size= 0.5) +
    geom_smooth(method = lm,  size = 0.3) + 
    ylim(c(0, 100)) + xlab("2020 vote share for the Democrats") + 
    xlim(c(0, 100)) + ylab("COVID-19 vaccination coverage") +
    scale_color_manual(values = mycols2) +
    theme_pubclean() +
    stat_cor(method = "pearson", aes(label = ..r.label..),  
             label.x = 0,  show.legend = FALSE, label.y.npc = "bottom",   size=4) + 
  theme(legend.position = "left") -> p1

  ggsave( here(paste("output/", 'FIG-S1A-', Sys.Date(), '.png', sep="")),
          device = png(),
          plot = p1,
          width =17,
          bg = "white",
          height = 14,
          units = "cm", 
          dpi = 300,
          limitsize = TRUE)
  dev.off()
  
  p1
  
```

\newpage

## SUPPLEMENTAL FIGURE S1 B

Comparison of correlation coefficient across state-level COVID-19 vaccination estimates by month. We have omitted Washington D.C. as an outlier.  Pearson correlation coefficient (R) shown by month in panels A,C,D, and as a bar plot in panel B.

```{r}

# CDC
 cv19_bymonth %>%
    mutate(month2 = month(month, label = TRUE, abbr = FALSE)) %>%
    filter(!is.na(vote_share) & date > ymd("2021-01-01") & date < ymd("2022-12-01")) %>%
    filter(state_code != "DC") %>%
    mutate(Year_Month =  format(date, '%Y-%m'))  %>%
    select(vote_share, Year_Month, state_code, est) %>%
    arrange(state_code, Year_Month) %>%
    group_by(state_code) %>%
    mutate(absincr = est - lag(est)) %>%
    mutate(relincr = est / lag(est)) %>%
    filter(Year_Month > "2021-01")   %>%
    group_by(Year_Month) %>%
    summarize(`Pearson correlation`=cor(vote_share , absincr)) -> corrtab
  
  corrtab %>%
   # mutate(order = 1/rank(Year_Month)) %>%
    mutate(x_place = if_else(`Pearson correlation` > 0, -0.4, 1)) %>%
    ggplot(aes(x=`Pearson correlation` , y=Year_Month, fill = Year_Month))  +
    geom_col() +
    theme_pubclean() +
    xlim(c(-0.4, 1)) + 
    xlab("Pearson correlation coefficient") + ylab("Month") + 
    geom_text(aes(label = round(`Pearson correlation`, 2), color = Year_Month, hjust = x_place)) +
    scale_fill_manual(values = mycols2[2:length(mycols2)] ) +
    scale_color_manual(values = mycols2[2:length(mycols2)] ) +
    theme(legend.position="none") -> p2
    
  ggsave( here(paste("output/", 'FIG-S1B-', Sys.Date(), '.png', sep="")),
          device = png(),
          plot = p2,
          width =17,
          bg = "white",
          height = 14,
          units = "cm", 
          dpi = 300,
          limitsize = TRUE)
  dev.off()
 
  p2
```


\newpage

## SUPPLEMENTAL FIGURE S1 C

Comparison of correlation coefficient across state-level COVID-19 vaccination estimates by month. We have omitted Washington D.C. as an outlier.  Pearson correlation coefficient (R) shown by month in panels A,C,D, and as a bar plot in panel B.

```{r}
# NIS-ACM  
    ## if there is an error (no plot), check if new data have been added in and amend code 
    ## coding of variables has changed over time (e.g. "All Adults 18+" vs "All adults 18+")
  statev %>%
    filter(indicator_category == "Vaccinated (≥1 dose)") %>%
    filter(time_type == "Monthly") %>%
    mutate(Year_Month =  format (ym(paste(year, match(month, month.name), sep = "")), '%Y-%m'))  %>%
    mutate(elect_year = 2020) %>%
    filter(state_code != "DC") %>%
    filter(group_name == "All adults 18+" |group_name == "All Adults 18+") %>%
    left_join(elect, by=c("state_code", "elect_year")) %>%
    filter(!is.na(vote_share)) %>%
    ggplot(aes(x=100*vote_share, y=est, color=Year_Month)) +
    geom_point(size= 0.5) +
    geom_smooth(method = lm,  size = 0.5) + 
    ylim(c(0, 100)) + xlab("2020 vote share for the Democrats") + 
    xlim(c(0, 100)) + ylab("COVID-19 vaccination coverage") +
    scale_color_manual(values = mycols2[5:length(mycols2)]) + # start so that color ordering follows the others
    theme_pubclean() +
    stat_cor(method = "pearson", aes(label = ..r.label..),  
             label.x = 0,  show.legend = FALSE, label.y.npc = "bottom",   size=4) + 
    guides(color = guide_legend(reverse=TRUE)) +
   theme(legend.position = "left") -> p3

  ggsave( here(paste("output/", 'FIG-1C-', Sys.Date(), '.png', sep="")),
          device = png(),
          plot = p3,
          width =17,
          bg = "white",
          height = 14,
          units = "cm", 
          dpi = 300,
          limitsize = TRUE)
  dev.off()
  
  p3
  
```

\newpage

- Obtain linear predictions by month for CDC surveillance data on COVID-19 vaccination

```{r}


 cv19_bymonth %>%
    mutate(month2 = month(month, label = TRUE, abbr = FALSE)) %>%
    filter(!is.na(vote_share) & date > ymd("2021-01-01") & date < ymd("2022-12-01")) %>%
    filter(state_code != "DC") %>%
    mutate(Year_Month =  format(date, '%Y-%m')) %>%
    mutate(vote_perc = 100*vote_share)  -> fits_cdc

print("CDC linear regression by month")
    lmList(est ~ vote_perc | Year_Month, data=fits_cdc)

```

\newpage

- Obtain linear predictions by month for NIS-ACM data on COVID-19 vaccination
    
```{r}

 statev %>%
    filter(indicator_category == "Vaccinated (≥1 dose)") %>%
    filter(time_type == "Monthly") %>%
    mutate(Year_Month =  format (ym(paste(year, match(month, month.name), sep = "")), '%Y-%m'))  %>%
    mutate(elect_year = 2020) %>%
    filter(state_code != "DC") %>%
    filter(group_name == "All adults 18+" |group_name == "All Adults 18+") %>%
   # mutate(group_category = as.factor(group_category)) %>%
    left_join(elect, by=c("state_code", "elect_year")) %>%
    mutate(vote_perc = 100*vote_share) -> fits_acm

print("ACM linear regression by month")
    lmList(est ~ vote_perc | Year_Month, data=fits_acm)

```

\newpage

## SUPPLEMENTAL FIGURE S2

Absolute monthly difference in COVID-19 vaccination coverage, compared at state-level to previous month. Shading represents 2020 vote share for the Democrats. Data: CDC Surveillance

```{r}

# CDC
# Supplemental figure
cv19_bymonth %>%
  mutate(month2 = month(month, label = TRUE, abbr = FALSE)) %>%
  filter(!is.na(vote_share) & date > ymd("2021-01-01") & date < ymd("2022-12-01")) %>%
  filter(state_code != "DC") %>%
  mutate(Year_Month =  format(date, '%Y-%m'))  %>%
  select(vote_share, Year_Month, state_code, est) %>%
  arrange(state_code, Year_Month) %>%
  group_by(state_code) %>%
  mutate(absincr = est - lag(est)) %>%
  filter(Year_Month > "2021-01")   %>%
  ggplot(aes(x=Year_Month , y=absincr, colour=100*vote_share, group = state_code))  +
  geom_line(size = 0.5) +
  ylim(c(-5, 30)) + ylab("Absolute monthly difference (%)") + 
  xlab("Month") + labs(colour = "Vote share (%)") +
  scale_color_gradient(low = "yellow",  high = "navyblue") +
  theme_pubclean() +
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   -> p5

  ggsave( here(paste("output/", 'FIG-S2-', Sys.Date(), '.png', sep="")),
          device = png(),
          plot = p5,
          width =17,
          bg = "white",
          height = 14,
          units = "cm", 
          dpi = 300,
          limitsize = TRUE)
  dev.off()
  
  p5
  
```
Footnote 
The data are presented as they were reported, and no data cleaning were done to adjust for potential reporting changes and/or errors; e.g. reductions in population level vaccination coverage observed for some states.

